---
title: "stat462"
output:
  pdf_document: default
  html_document: default
  word_document: default
date: "2026-02-18"
---

```{r}
library(ggplot2)
library(tidyr)
library(tidyverse)
```

# Summary Dataset:
```{r }


diabetes <- read.csv(file="~/Desktop/diabetes2.csv",header=T)

summary(diabetes)
```

# Box Plot:
```{r }
par(mfrow = c(2, 2))
par(mar = c(3, 3, 2, 1)) 

for (i in 1:8) {
  boxplot(diabetes[, i] ~ diabetes$Outcome, 
          main = names(diabetes)[i],
          xlab = names(diabetes)[i],
          ylab = "Outcome", 
          col = c("green", "yellow"),
          names = c("No Diabetes", "Diabetes"))
}
```

# Histogram:
```{r }
par(mfrow=c(3,3))

hist(diabetes$Pregnancies, main = "Histogram of Pregnancies", 
     xlab = "Pregnancies", col = "skyblue", border = "darkblue")

hist(diabetes$Glucose, main = "Histogram of Glucose", 
     xlab = "Glucose", col = "skyblue", border = "darkblue")

hist(diabetes$BloodPressure, main = "Histogram of BloodPressure", 
     xlab = "BloodPressure", col = "skyblue", border = "darkblue")

hist(diabetes$SkinThickness, main = "Histogram of SkinThickness", 
     xlab = "SkinThickness", col = "skyblue", border = "darkblue")

hist(diabetes$Insulin, main = "Histogram of Insulin", 
     xlab = "Insulin", col = "skyblue", border = "darkblue")

hist(diabetes$BMI, main = "Histogram of BMI", 
     xlab = "BMI", col = "skyblue", border = "darkblue")

hist(diabetes$DiabetesPedigreeFunction, main = "Histogram of DiabetesPedigreeFunction", 
     xlab = "DiabetesPedigreeFunction", col = "skyblue", border = "darkblue")

hist(diabetes$Age, main = "Histogram of Age", 
     xlab = "Age", col = "skyblue", border = "darkblue")

hist(diabetes$Outcome, main = "Histogram of Outcome", 
     xlab = "Outcome", col = "skyblue", border = "darkblue")
```

# Non-logical Values:
```{r}
# Finding missing values at first:
colSums(is.na(diabetes))

#Count 0 in each column:
sapply(diabetes, function(x) sum(x == 0, na.rm = TRUE))

#Count clinically nonsensical 0 only
nonsense_zero_counts <- c(
  Glucose = sum(diabetes$Glucose == 0, na.rm = TRUE),
  BloodPressure = sum(diabetes$BloodPressure == 0, na.rm = TRUE),
  SkinThickness = sum(diabetes$SkinThickness == 0, na.rm = TRUE),
  Insulin = sum(diabetes$Insulin == 0, na.rm = TRUE),
  BMI = sum(diabetes$BMI == 0, na.rm = TRUE)
)

nonsense_zero_counts
```

# Recode Selected 0 Values to NA:
```{r}
#Recode non-logical 0 values into NA:
diabetes_clean <- diabetes

diabetes_clean$Glucose[diabetes_clean$Glucose == 0] <- NA
diabetes_clean$BloodPressure[diabetes_clean$BloodPressure == 0] <- NA
diabetes_clean$SkinThickness[diabetes_clean$SkinThickness == 0] <- NA
diabetes_clean$Insulin[diabetes_clean$Insulin == 0] <- NA
diabetes_clean$BMI[diabetes_clean$BMI == 0] <- NA

#Check NA counts after recoding:
colSums(is.na(diabetes_clean))
```

# % of Missing Value:
```{r}
# % of Missing value after recoding:
missing_count <- colSums(is.na(diabetes_clean))
missing_percentage <- round((missing_count / nrow(diabetes_clean)) * 100, 2)

# Sumummary table of missing value:
missing_summary <- data.frame(
  Variable = names(missing_count),
  Missing_Count = missing_count,
  Missing_Percentage = missing_percentage
)

missing_summary
```

# Median Calculation & Imputation:
```{r}
#Median values for variables with non-logical 0 values:
glucose_median <- median(diabetes_clean$Glucose, na.rm = TRUE)
bloodpressure_median <- median(diabetes_clean$BloodPressure, na.rm = TRUE)
skinthickness_median <- median(diabetes_clean$SkinThickness, na.rm = TRUE)
insulin_median <- median(diabetes_clean$Insulin, na.rm = TRUE)
bmi_median <- median(diabetes_clean$BMI, na.rm = TRUE)

#Show median values:
glucose_median
bloodpressure_median
skinthickness_median
insulin_median
bmi_median

#Median imputation for NA values:
diabetes_imputed <- diabetes_clean

diabetes_imputed$Glucose[is.na(diabetes_imputed$Glucose)] <- glucose_median
diabetes_imputed$BloodPressure[is.na(diabetes_imputed$BloodPressure)] <- bloodpressure_median
diabetes_imputed$SkinThickness[is.na(diabetes_imputed$SkinThickness)] <- skinthickness_median
diabetes_imputed$Insulin[is.na(diabetes_imputed$Insulin)] <- insulin_median
diabetes_imputed$BMI[is.na(diabetes_imputed$BMI)] <- bmi_median
```
# Immputed dataset:
```{r}
# Check the imputed dataset:
head(diabetes_imputed)
summary(diabetes_imputed)

# Confirm that no missing values remain:
colSums(is.na(diabetes_imputed))

# Export the final cleaned and imputed dataset:
write.csv(diabetes_imputed,
          "diabetes_imputed.csv",
          row.names = FALSE)
```
#visualization of historgrams
```{r}
library(ggplot2)
library(tidyr)
library(dplyr)

diabetes_long <- diabetes %>%
  pivot_longer(cols = -Outcome, names_to = "Variable", values_to = "Value")

ggplot(diabetes_long, aes(x = Value, fill = factor(Outcome))) +
  geom_histogram(alpha = 0.6, position = "identity", bins = 30) +
  facet_wrap(~ Variable, scales = "free", ncol = 4) +
  scale_fill_manual(values = c("skyblue", "lightcoral"),
                    labels = c("No Diabetes", "Diabetes")) +
  labs(title = "Distribution of Variables by Diabetes Outcome",
       x = "Value", y = "Count", fill = "Outcome") +
  theme_minimal() +
  theme(legend.position = "bottom")

```

#visualization of violin plots
```{r}
diabetes_imputed_long <- diabetes_imputed %>%
  pivot_longer(cols = -Outcome, names_to = "Variable", values_to = "Value")

ggplot(diabetes_imputed_long, aes(x = Outcome, y = Value, fill = Outcome)) +
  geom_violin(trim = FALSE, alpha = 0.6) +
  geom_boxplot(width = 0.15, fill = "white", alpha = 0.8) +
  facet_wrap(~ Variable, scales = "free_y") +
  labs(x = "Outcome", y = "Value",
       title = "Violin Plot of Variables by Diabetes Outcome") +
  scale_fill_manual(values = c("green", "red")) +
  theme_minimal()
```

# Correlation Matrix:
```{r}
library(corrplot)

#Select only predictor variables only
core_vars <- c("Pregnancies", "Glucose", "BloodPressure", "SkinThickness",
               "Insulin", "BMI", "DiabetesPedigreeFunction", "Age")

cat("x:", core_vars, "\n")
cat("y:", "Outcome", "\n")

#Correlation matrix using imputed data
cor_mat <- cor(diabetes_imputed[, core_vars], use = "pairwise.complete.obs")

# Correlation matrix values:
round(cor_mat, 2)

# Visualize correlation plot:
corrplot(cor_mat,
         method = "color",
         type = "lower",
         diag = FALSE,
         addCoef.col = "black",
         tl.col = "black",
         tl.cex = 0.8,
         number.cex = 0.7,
         col = colorRampPalette(c("#2166AC", "white", "#D6604D"))(200),
         main = "Correlation Plot of Diabetes Variables",
         mar = c(0, 0, 2, 0))
```

# Multicollinearity:
# 1. Core Model (exclude Isulin & Skin Thickness):
```{r}
library(car)

# Fit logistic regression model using the imputed dataset (without Isulin & Skin Thickness due to overlapping)
model_core <- glm(Outcome ~ Pregnancies + Glucose + BloodPressure + BMI +
                    DiabetesPedigreeFunction + Age,
                  data = diabetes_imputed,
                  family = binomial)

# Show model summary
summary(model_core)

# Calculate Variance Inflation Factor (VIF)
vif(model_core)
```

# 2. Full Model:
```{r}
# Full model including SkinThickness and Insulin:
model_full <- glm(Outcome ~ Pregnancies + Glucose + BloodPressure +
                    SkinThickness + Insulin + BMI +
                    DiabetesPedigreeFunction + Age,
                  data = diabetes_imputed,
                  family = binomial)

# Show model summary
summary(model_full)

# Calculate VIF for the full model
vif(model_full)
```

The VIF results showed that multicollinearity was low in both models, so overlap between variables was not a major issue.

So, we did not leave out `SkinThickness` and `Insulin` because of multicollinearity. They were excluded from the main model because they had many unrealistic 0 values, needed a lot of imputation, and did not improve the model. Therefore, the core model (without `SkinThickness` and `Insulin`) was used as the main training model, while the full model was only kept for comparison.

# Train-test data splitting:
```{r}
# Partition the dataset
library(caret)

set.seed(7835)

# Convert outcome to factor for classification tree
diabetes_imputed$Outcome <- factor(diabetes_imputed$Outcome,
                                   levels = c(0, 1),
                                   labels = c("NoDiabetes", "Diabetes"))

# Stratified split to preserve class balance
train_index <- createDataPartition(diabetes_imputed$Outcome,
                                   p = 0.8,
                                   list = FALSE)

train <- diabetes_imputed[train_index, ]
test  <- diabetes_imputed[-train_index, ]

# Check split sizes
nrow(train)
nrow(test)

# Check class distribution
prop.table(table(train$Outcome))
prop.table(table(test$Outcome))
```

#Setup for selection
```{r}
#original model with 8 variables 
full_model <- glm(Outcome ~ ., data = train, family = binomial)

#empty model for later selections 
null_model <- glm(Outcome ~ 1, data = train, family = binomial)


summary(full_model)
summary(null_model)
```



#forward selection
```{r}
forward_model <- step(
  null_model,                    
  scope = list(lower = null_model, upper = full_model), 
  direction = "forward",        
  trace = 0 #show with 1, hide with 0                     
)

summary(forward_model)

```


#backward elimination
```{r}
backward_model <- step(
  full_model,
  direction = "backward",        
  trace = 0
)


summary(backward_model)
```


#Stepwise selection 
```{r}
stepwise_model <- step(
  null_model,                   
  scope = list(lower = null_model, upper = full_model),
  direction = "both",           
  trace = 0
)


summary(stepwise_model)
```

## Odds Ratio (OR) calculations with table 
```{r}
summary(stepwise_model)

exp_coef <- exp(coef(stepwise_model))
exp_ci <- exp(confint(stepwise_model))


coef_table <- data.frame(
  Variable = names(coef(stepwise_model)),
  Coefficient = round(coef(stepwise_model), 4),
  Odds_Ratio = round(exp_coef, 3),
  CI_lower = round(exp_ci[, 1], 3),
  CI_upper = round(exp_ci[, 2], 3)
)

print(coef_table)
```
#Once glucose increase 1mg/dL, the risk of diabetese increase 3.9% (explnantion above) All predictors except age and blood pressure had 95% confidence intervals that did not contain 1, confirming their statistical significance. 


#Choose the stepwise selection and test 
```{r}
pred_prob <- predict(stepwise_model, 
                     newdata = test, 
                     type = "response")

pred_class <- ifelse(pred_prob > 0.5, "Diabetes", "NoDiabetes")
pred_class <- factor(pred_class, levels = c("NoDiabetes", "Diabetes"))

accuracy <- mean(pred_class == test$Outcome)
cat("Accuracy:", round(accuracy, 4), "\n")

conf_matrix <- table(Predicted = pred_class, Actual = test$Outcome)
print(conf_matrix)

sensitivity <- conf_matrix["Diabetes", "Diabetes"] / sum(conf_matrix[, "Diabetes"])
cat("Sensitivity:", round(sensitivity, 4), "\n")

specificity <- conf_matrix["NoDiabetes", "NoDiabetes"] / sum(conf_matrix[, "NoDiabetes"])
cat("Specificity:", round(specificity, 4), "\n")


```


# AUC model for LR
```{r}
library(pROC)
test_outcome_num <- ifelse(test$Outcome == "Diabetes", 1, 0)

roc_obj <- roc(test_outcome_num, pred_prob)
auc_value <- auc(roc_obj)
cat("AUC:", round(auc_value, 4), "\n")

plot(roc_obj, col = "blue", lwd = 2, 
     main = paste("ROC Curve (AUC =", round(auc_value, 3), ")"))
```

# Fit a classification tree:
```{r}
# Classification tree:
library(rpart)
library(rpart.plot)

# Specify target and input variables
formula <- Outcome ~ .

# Fit classification tree on training set
dtree_diabetes <- rpart(formula,
                        data = train,
                        method = "class")

# Print the tree
print(dtree_diabetes)

# Show complexity parameter table
printcp(dtree_diabetes)
```

# Visualize the decision tree
```{r}
# Plot the classification tree
rpart.plot(dtree_diabetes,
           type = 3,
           fallen.leaves = FALSE,
           extra = 104,
           box.palette = "GnBu",
           branch.lty = 1,
           shadow.col = "gray",
           nn = TRUE,
           main = "Classification Tree for Diabetes Prediction")
```

# Check variable importance
```{r}
# Variable importance
dtree_diabetes$variable.importance
```

# Predictions on train and test sets
```{r}
# Predict on training and test sets
train$predicted_tree <- predict(dtree_diabetes,
                                train,
                                type = "class")

test$predicted_tree <- predict(dtree_diabetes,
                               test,
                               type = "class")

# Confusion matrix on test set
tree_cm <- confusionMatrix(test$predicted_tree,
                           test$Outcome,
                           positive = "Diabetes")

tree_cm

# Main metrics for original tree
tree_accuracy <- tree_cm$overall["Accuracy"]
tree_sensitivity <- tree_cm$byClass["Sensitivity"]
tree_specificity <- tree_cm$byClass["Specificity"]

tree_accuracy
tree_sensitivity
tree_specificity

# False positives and false negatives
tree_table <- table(Predicted = test$predicted_tree,
                    Actual = test$Outcome)

tree_table

tree_false_positives <- tree_table["Diabetes", "NoDiabetes"]
tree_false_negatives <- tree_table["NoDiabetes", "Diabetes"]

tree_false_positives
tree_false_negatives
```

# Prune the tree:
```{r}
# Find best CP value using lowest xerror
best_cp <- dtree_diabetes$cptable[which.min(dtree_diabetes$cptable[, "xerror"]), "CP"]
best_cp

# Prune the tree
pruned_dtree_diabetes <- prune(dtree_diabetes, cp = best_cp)

# Plot pruned tree
rpart.plot(pruned_dtree_diabetes,
           type = 3,
           fallen.leaves = FALSE,
           extra = 104,
           box.palette = "GnBu",
           branch.lty = 1,
           shadow.col = "gray",
           nn = TRUE,
           main = "Pruned Classification Tree for Diabetes Prediction")
```

# Evaluate pruned tree:
```{r}
# Predict with pruned tree
test$predicted_pruned <- predict(pruned_dtree_diabetes,
                                 test,
                                 type = "class")

# Confusion matrix for pruned tree
pruned_cm <- confusionMatrix(test$predicted_pruned,
                             test$Outcome,
                             positive = "Diabetes")

pruned_cm

# Main metrics for pruned tree
pruned_accuracy <- pruned_cm$overall["Accuracy"]
pruned_sensitivity <- pruned_cm$byClass["Sensitivity"]
pruned_specificity <- pruned_cm$byClass["Specificity"]

pruned_accuracy
pruned_sensitivity
pruned_specificity
```

# Compare orginal tree vs pruned tree:
```{r}
# Summary comparison
tree_comparison <- data.frame(
  Model = c("Original Decision Tree", "Pruned Decision Tree"),
  Accuracy = c(tree_accuracy, pruned_accuracy),
  Sensitivity = c(tree_sensitivity, pruned_sensitivity),
  Specificity = c(tree_specificity, pruned_specificity)
)

tree_comparison
```



```{r}
library(pROC)


tree_prob <- predict(dtree_diabetes, newdata = test, type = "prob")
test_diabetes_prob <- tree_prob[, "Diabetes"]
test_outcome_num <- ifelse(test$Outcome == "Diabetes", 1, 0)

tree_roc <- roc(test_outcome_num, test_diabetes_prob)
tree_auc <- auc(tree_roc)
cat("Original Decision Tree AUC:", round(tree_auc, 4), "\n")


pruned_prob <- predict(pruned_dtree_diabetes, newdata = test, type = "prob")
pruned_diabetes_prob <- pruned_prob[, "Diabetes"]


pruned_roc <- roc(test_outcome_num, pruned_diabetes_prob)
pruned_auc <- auc(pruned_roc)
cat("Pruned Decision Tree AUC:", round(pruned_auc, 4), "\n")


plot(tree_roc, col = "blue", lwd = 2, 
     main = "ROC Curves: Decision Trees vs Logistic Regression")
lines(pruned_roc, col = "red", lwd = 2)
lines(roc_obj, col = "darkgreen", lwd = 2) 


legend("bottomright",
       legend = c(paste("Original Tree (AUC =", round(tree_auc, 3), ")"),
                  paste("Pruned Tree (AUC =", round(pruned_auc, 3), ")"),
                  paste("Logistic Regression (AUC =", round(auc_value, 3), ")")),
       col = c("blue", "red", "darkgreen"),
       lwd = 2)
```





